Phylogenetics and the human microbiome.

### 2014-04-11 reviews/response.txt

> 15-xx. This seems a strong point of contact between metazoan
> phylogenetics and the microbiome folks. UniFrac is clearly described
> on its own terms, except perhaps where I would have written "with the
> fraction of edges connecting samples that are common to two samples"
> rather than "edges that connect between samples" (which doesn't
> strictly make sense). I believe there is the same measure of
> beta-diversity in the community-phylogenetics literature? What is the
> closest analogue?

> 15-4955. The description of Evans and Matsen seems a bit detailed in
> the context -

> 16-16. This requires another sentence of explanation. From this
> reading, the Unifrac method offers a single distance between a pair of
> samples (a ratio of two numbers of edges), rather than a matrix.

I have rephrased so that it's clear that the matrix comes from applying
the measure to all pairs of sequences in a sample.

@@

> 16-42..51. This is a short paragraph that would seem to explicitly
> link HMR and SB. Why is it so telegraphic?

I have added a clarifying phrase, but I'm not sure what more is needed.
The two communities have both identified a given technique (although
neither has really gotten much traction).

@@

> 17-45. This seems an odd heading for this section, which seems to be
> about horizontal gene transfer. Given the amount of technical detail
> elsewhere, why are we not told how horizontal gene transfer is
> inferred in these HMR studies?

Explained now, thanks.

> Phylogenetic inference. I found this survey very useful. Is there a
> reason this is not an earlier section of the paper, given these are
> the trees upon which many of the comparisons referred to in previous
> sections are based?

I suppose I wanted to talk about all of the interesting things that
microbiome researchers are doing with trees before
borderline-criticizing the ways that they make them! The readers of SB
know what it means to infer a tree, so I don't think that this section
is a prerequisite for the previous ones.

@@

> 19-24. "which has ported in the" Is this the correct structure?

That was what was intended, but I've updated it to use less
programmer-speak.

@@

> 20-50. "different parameter regimes" - this phrase is used several
> times in this review. What does it mean here? When would we prefer to
> add sequences to a fixed tree rather than re-infer the totality? Is
> there an analogous situation in SB - i.e. is there something SB can
> learn from HMR here?

I've simply removed the sentence. The idea of this paragraph is to
motivate updating of phylogenetic trees with new sequences, and I've
more fully explained that and added a reference.

@@

> I found the challenges section to be rich and well-written, with
> questions that will resonate with the SB community - e.g. the
> discussion of the proxy problem (how edge lengths predict feature
> diversity). The fact that 16S distance does predict whole genome
> distance is highlighted here - it is presented in previous sections,
> but perhaps it could be given more prominence there, given its
> prominence here. This is a general problem.

I have added a sentence emphasizing that this assumption underlies the
way researchers interpret UniFrac.

@@

> 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
> ecology" to remind us that HMR is the focus here?

Added, thanks.

> 23-23. "Whole organism approach"  Will this approach be known to SB
> readers? I cannot imagine what is being done here.

You are right that this was not explained properly. I have simply
removed the paragraph.

@@

> 25-31. "At least for the next several years we can expect more related
> work." What does this mean? Related to what has been reported above?

I have made this more specific.

> 25-32. The future

> I found the final sentence to leave us dangling. After having reviewed
> this manuscript, I could not say exactly where the intersection of SB
> and HMR will be most fruitful for HMR, nor what in HMR might be
> fruitful for SB. The conclusion seems to suggest that phylogenetics
> would not be used in diagnostics, for example.

I have clarified this sentence now to clarify that some applications
will go but many new ones will be needed.

> Will it be important in primary research then? In constructing the
> taxonomies the diagnostics may require?

The rest of the paper is about how phylogenetics has impacted HMR, and
these applications will continue into the future. I'm not sure how to
make this more clear.

> Finally, the abstract says the review ends by describing
> "the current challenges to phylogenetics coming from this type of
> work."  My reading of the relevant sections is that there is more
> discussion of challenges to phylogenetic HMR research (sampling
> biases, robustness of UniFrac to incorrect tree reconstruction) than
> to phylogenetics per se. The discussion of how to update known trees
> with new sequences is an exception.

I suppose I don't differentiate between the two.

@@

> 4. On page 6, line 56, maybe also specify that "survey" is the
> equivalent of barcoding, a term with which the Systematic Biology's
> community is more familiar.

Added, thanks.

> 5. On page 14, lines 14-19, maybe mention the widespread use of
> "rarefaction curves" to enable the comparison between samples of
> different sequencing depth or methods (e.g. 454 vs. Illumina).

I do describe strategies for read count normalization in the next paragraph, but have added the "rarefaction curve" terminology.

@@

> 6. On page 17, lines 54-55, please add some recent references on the
> current debate around the rooting of the placental tree. I don't think
> that most readers of Syst. Biol. will be familiar with this
> discussion.

Good point. I've put in the two 2013 MBE papers. If there is something else I should cite let me know.

### 2014-04-11 sbreview.tex
WIP

\section{The human microbiome}
The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
In this section I will briefly review what is known about the human microbiome and its effect on our health.

@@

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

@@

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is in a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16S'') as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, that it can be calculated in linear time in contrast to many applications of the earth-movers distance, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

@@

The most common way to use a between-sample pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
\citep[e.g.][]{lozupone2007global,costello2009bacterial,yatsunenko2012human}, while the lack of such a difference is interpreted as showing that the communities are not different overall.

@@

In another vein, \citet{la2012statistical} consider the induced taxonomic tree of a sample as a statistical object and, using a framework where a sampling probability is defined in terms of a distance between such induced trees, define and investigate maximum likelihood estimation of and likelihood ratio tests for these trees.
They focus on distances between trees induced by matrix metrics on the corresponding adjacency matrices.
A similar framework was used by \citet{steel2008maximum} to construct maximum likelihood supertrees.

@@

16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in 16S ancestry.

@@

\subsection{Genome-scale inquiries using phylogenetics}
With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
As described above, whole-genome data sets are typically used to directly infer functional information rather than information concerning ancestry.
The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.

Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.

@@

\subsection{Alignment and tree inference}
In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
Furthermore, given that the underlying data sets are typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
Additionally, as specified below, these data sets can be very large.
There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.

@@

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut program \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

The scale of the data has motivated strategies other than complete phylogenetic inference, such as the insertion of sequences into an existing phylogenetic tree.
Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was the parsimony insertion tool in the ARB program by \citet{ludwig2004arb}.
ARB is commonly used to reconstruct a full tree by direct insertion.

@@

% Networks are not considered.

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
For example, sequence data sets are continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.

@@

The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
If we allow an arbitrary ``hidden'' object, what such object would perform best?
For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[see][for interesting recent results]{szollHosi2013efficient,szollHosi2013lateral} could be used.

@@

An important project for microbial ecology is to model community assembly, and perhaps further phylogeny-aware methods could be used.
One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.

@@

The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \citep{borenstein2008large,greenblum2012metagenomic}.
A clear limitation to this approach is that cellular boundaries are ignored: populations are not a freely diffusing soup.
Could these approaches be improved by using phylogenetic methods to reconstruct the compartmentalization of genomes and processes into cells?
% As an extension microbes have to be close to one another-- requires population modeling.

@@

In general, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.

Another type of normalization seeks to infer the true abundances from noisy observations of the various taxonomic groups or OTUs.
\citet{holmes2012dirichlet} and \citet{la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
\citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian model for read counts.
This work could be extended to a phylogenetic context by making use of the relationship between OTUs, and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
For example, routine 16S sequencing is likely to be replaced soon by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for assignment of a single microbe grown in culture to a database entry \citep{clark2013matrix}.
However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[reviewed in][]{Rogers2013271}.
Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
For all of these measures, it will be important to have rigorous means of quantifying uncertainty for robust diagnostic applications.

>>>>

\section{The human microbiome}
The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
Our microbiomes have been historically transmitted vertically between mother and infant by seeding in the birth canal and breastfeeding;  immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
In this section I will briefly review what is known about the human microbiome and its effect on our health.

@@

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \cite{Findley2013-rm}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

@@

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is in a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16S'') as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled detection of almost all bacteria and archaea regardless of whether they can be cultured.
In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys,'' which is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

Faith2009-gs

@@

The most common way to use a distance matrix obtained from applying UniFrac to all pairs of samples is to apply an ordination method such as principal coordinates analysis.
Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
\citep[e.g.][]{lozupone2007global,costello2009bacterial,yatsunenko2012human}, while the lack of such a difference is interpreted as showing that the communities are not different overall.

@@

In another vein, \citet{la2012statistical} consider the induced taxonomic tree of a sample as a statistical object and, using a framework where a sampling probability is defined in terms of a distance between such induced trees, define and investigate maximum likelihood estimation of and likelihood ratio tests for these trees.
They focus on distances between trees induced by matrix metrics on the corresponding adjacency matrices.
A similar framework was used by \citet{steel2008maximum} to construct maximum likelihood supertrees, for which they use common distance measures in phylogenetics such as the subtree-prune-regraft metric.

@@

16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
Indeed, researchers using UniFrac frequently don't think of their comparisons as being in terms of a single gene, but rather in terms of an abstracted measure of community function.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in 16S ancestry.

@@

\subsection{Horizontal gene transfer}
With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
As described above, whole-genome data sets are typically used to directly infer functional information rather than information concerning ancestry.
The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.

Nevertheless, the work that has been done concerning horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
Horizontal transfer of genes is inferred in these studies by finding highly similar subsequences in otherwise less related organisms.

@@

\subsection{Alignment and tree inference}
In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
Furthermore, given that the underlying data sets are typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
Additionally, as specified below, these data sets can be very large.
There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree \citep{morgan2013heterogeneous,romiguier2013less}.
Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.

@@

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \cite{Stamatakis2006-yz}.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

The scale of the data has motivated strategies other than complete phylogenetic inference, such as the insertion of sequences into an existing phylogenetic tree.
Although such insertion has long been used as a means to build a phylogenetic tree sequentially \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was the parsimony insertion tool in the ARB program by \citet{ludwig2004arb}.
ARB is commonly used to reconstruct a full tree by direct insertion.

@@

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed, with the idea that such an algorithm would retain the efficiency characteristics of placement algorithms while allowing the reference tree to change.
For example, sequence data sets are continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
\citet{Izquierdo-Carrasco2014-hu} have taken a step in this direction by developing an informatic framework that updates alignments and builds larger trees using previous smaller trees as starting points.

@@

The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
If we allow an arbitrary ``hidden'' object, what such object would perform best?
An important first step in this direction has been taken by \citet{Parks2012-os} who define community comparison metrics on phylogenetic split systems.
An alternative would be to use collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[e.g.][]{szollHosi2013efficient,szollHosi2013lateral}.

@@

Developing community assembly models forms an important project for microbial ecology generally and human microbial ecology in particular, phylogeny-aware methods could be further developed.
One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.

@@

@@

For both alpha and beta diversity studies, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.

Another type of normalization seeks to infer the true abundances from noisy observations of the various taxonomic groups or OTUs.
\citet{holmes2012dirichlet} and \citet{la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
\citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian mixture model for read counts, while \citet{McMurdie2014-mn} claim better performance using a Gamma-Poisson mixture.

This work could be extended to a phylogenetic context by making use of the relationship between OTUs, and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect the research questions described above to continue to unfold.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Opportunities for clinical applications will present themselves, although the specifics will change.
For example, routine 16S sequencing is likely to be replaced soon by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for assignment of a single microbe grown in culture to a database entry \citep{clark2013matrix}.
However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods (possibly including phylogenetics) will still be required \citep[reviewed in][]{Rogers2013271}.
Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}, and this genome-scale data requires evolutionary analysis methods to interpret it.
For all of these measures, it will be important to have rigorous means of quantifying uncertainty for robust diagnostic applications.
